Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
minor edits

The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

To yield functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015]. 
We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits–particularly leaf hydraulic traits–predict Rt. We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors. We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].

@@

Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected nearby the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.

Crown position ($CP$)--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999. While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible). An analysis of crown position relative to height (Fig. 2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions (Fig. S3), with average tree height growth confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018. However, dominant and co-dominant trees were similar in height (Figs. 2d, S3). While some canopy positions undoubtedly shifted in the 52 years between the 1966 drought and our observations in 2018, the resulting bias would be more likely result in type II error with regards to testing the hypothesis that dominant trees had lower resistance, meaning that our test is conservative.

>>>>

The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

To yield functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in @bennett_larger_2015. 
We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors. We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].

@@

Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019. Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected nearby the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.

Crown position ($CP$)--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999. While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible). An analysis of crown position relative to height (Fig. 2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions (Fig. S3), with average tree height growth confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018. However, dominant and co-dominant trees were similar in height (Figs. 2d, S3).
